---
title: "simmr_ffvb vignette for rcpp testing"
author: "Emma Govan"
date: "7/18/2022"
output:
  html_document: default
  pdf_document: default
---


```{r,eval=FALSE}
library(simmr)
```

simmr_ffvb function for comparison
```{r}
simmr_ffvb<-function(simmr_in,
                     prior_control = list(
                       means = rep(
                         1,
                         simmr_in$n_sources),
                       sd = rep(
                         1,
                         simmr_in$n_sources),
                       c_0 = rep(
                         1,
                         simmr_in$n_tracers),
                       d_0 = rep(
                         1,
                         simmr_in$n_tracers))
){
  
  # Throw a warning if less than 4 observations in a group - 1 is ok as it wil do a solo run
  if (min(table(simmr_in$group)) > 1 & min(table(simmr_in$group)) < 4) warning("At least 1 group has less than 4 observations - either put each observation in an individual group or use informative prior information")
  
  
  output <- vector("list", length = simmr_in$n_groups)
  names(output) <- levels(simmr_in$group)
  K<-simmr_in$n_sources
  n_tracers <- simmr_in$n_tracers
  lambdares<-matrix(rep(NA, ((( K + (K * (K + 1)) / 2)) + n_tracers * 2) * simmr_in$n_groups),
                    nrow =((( K + (K * (K + 1)) / 2)) + n_tracers * 2),
                    ncol = simmr_in$n_groups)
  
  
  # Loop through all the groups
  for (i in 1:simmr_in$n_groups) {
    if (simmr_in$n_groups > 1) cat(paste("\nRunning for group", levels(simmr_in$group)[i], "\n\n"))
    
    curr_rows <- which(simmr_in$group_int == i)
    curr_mix <- simmr_in$mixtures[curr_rows, , drop = FALSE]
    
    # Determine if a single observation or not
    if (nrow(curr_mix) == 1) {
      cat("Only 1 mixture value, performing a simmr solo run...\n")
      solo <- TRUE
    } else {
      solo <- FALSE
    }
    
    
    n_tracers <- simmr_in$n_tracers
    n_sources <- simmr_in$n_sources
    s_names <- simmr_in$source_names
    K<-simmr_in$n_sources
    S <- 100
    source_means = simmr_in$source_means
    source_sds = simmr_in$source_sds
    correction_means = simmr_in$correction_means
    correction_sds = simmr_in$correction_sds
    concentration_means = simmr_in$concentration_means
    y = curr_mix
    
    
    
    
    # Source in all the generic functions
    
    #source("FF_VB_generic_functions_correct.R")
    
    # K <- 3; lambda <- c(rnorm(K+K*(K+1)/2), rgamma(n_tracers*2, 1,1))
    # theta <- sim_theta(50, lambda)
    
    # Log of likelihood added to prior
    h = function(theta) {
      p <- exp(theta[1:K]) / (sum((exp(theta[1:K])))) #slightly hacky way of doing 1 or
      return(ifelse((simmr_in$n_tracers == 2), 
                    (sum(dnorm(y[, 1],
                        mean = sum(p * concentration_means[,1] * (source_means[, 1]+correction_means[,1]))/sum(p*concentration_means[,1]),
                        sd = sqrt(sum(p^2 * concentration_means[,1]^2 * (source_sds[, 1]^2+correction_sds[,1]^2))/sum(p^2*concentration_means[,1]^2) + 1 / theta[K + 1]),
                        log = TRUE)) +
                    sum(dnorm(y[, 2],
                      mean = sum(p * concentration_means[,2] * (source_means[, 2]+correction_means[,2]))/sum(p*concentration_means[,2]),
                      sd = sqrt(sum(p^2 * concentration_means[,2]^2 * (source_sds[, 2]^2+correction_sds[,2]^2))/sum(p^2*concentration_means[,2]^2) + 1 / theta[K + 2]),
                      log = TRUE))), 
                    (sum(dnorm(y,
                              mean = sum(p * concentration_means * (source_means+correction_means))/sum(p*concentration_means),
                              sd = sqrt(sum(p^2 * concentration_means^2 * (source_sds[, 1]^2+correction_sds^2))/sum(p^2*concentration_means^2) + 1 / theta[K + 1]),
                              log = TRUE
        )))) +
        sum(dnorm(theta[1:K], prior_control$means, prior_control$sd, log = TRUE)) +
        sum(dgamma(theta[(K + 1):(K + n_tracers)], shape = prior_control$c_0, rate = prior_control$d_0, log = TRUE)))
    }
    # h(theta[1,])
    
    log_q = function(lambda, theta) {
      mean <- lambda[1:K]
      # K*(K-1) precision terms
      chol_prec <- matrix(0, nrow = K, ncol = K)
      chol_prec[upper.tri(chol_prec, diag = TRUE)] <- lambda[(K + 1):(K + (K * (K + 1)) / 2)]
      # chol_prec[1,3] <- 0
      
      # This is a placeholder for more advanced code using chol_prec directly
      # prec <- crossprod(chol_prec)
      p1 <- matrix(theta[1:K] - mean, nrow = 1) %*% t(chol_prec)
      # log_det <- unlist(determinant(prec, logarithm = TRUE))["modulus"]
      return(-0.5 * K * log(2 * pi) - 0.5 * sum(log(diag(chol_prec))) - 0.5 * p1%*%t(p1)
             + sum(dgamma(theta[(K + 1):(K + n_tracers)],
                          shape = lambda[((K + (K * (K + 1)) / 2) + 1):(((K + (K * (K + 1)) / 2)) + n_tracers)],
                          rate = lambda[(((K + (K * (K + 1)) / 2)) + n_tracers + 1):(((K + (K * (K + 1)) / 2)) + n_tracers * 2)],
                          log = TRUE
             )))
    }
    
    sim_theta = function(S, lambda) {
      
      # For K parameters you will have
      # lambda is of length K+K*(K+1)/2 +n_tracers*2
      # mean <- lambda[1:K]
      # chol_prec is made up of lambda[(K + 1):(K+(K*(K+1))/2)]
      # Tau is made up of lambda[((K+(K*(K+1))/2)+1):((K+(K*(K+1))/2)+n_tracers*2)]
      # (f) ~ MVN(lambda[1:K], solve(crossprod(chol_prec)))
      
      
      mean <- lambda[1:K]
      # K*(K-1) precision terms
      chol_prec <- matrix(0, nrow = K, ncol = K)
      chol_prec[upper.tri(chol_prec, diag = TRUE)] <- lambda[(K + 1):(K + (K * (K + 1)) / 2)]
      
      # This is a placeholder for more advanced code using chol_prec directly
      theta <- cbind(
        t(rMVNormC(S, mu = mean, U = chol_prec)),
        matrix(rgamma(S * n_tracers,
                      shape = lambda[((K + (K * (K + 1)) / 2) + 1):(((K + (K * (K + 1)) / 2)) + n_tracers)],
                      rate = lambda[(((K + (K * (K + 1)) / 2)) + n_tracers + 1):(((K + (K * (K + 1)) / 2)) + n_tracers * 2)]
        ),
        nrow = S,
        ncol = n_tracers,
        byrow = TRUE
        )
      )
      
      return(theta)
    }
    
    # This contains all the generic functions for FF VB
    
    # Function to estimate different between joint and variational approx
    h_lambda <- function(lambda, theta, y) {
      return(h(theta) - log_q(lambda, theta))
    }
    # h_lambda(lambda, theta[1,], y)
    
    # Nable LB is the mean of delta_lqlt element-wise multiplied by h_lambda
    nabla_LB <- function(lambda, theta, c = rep(0, length(lambda))) {
      big_delta_lqlt <- t(apply(theta, 1, delta_lqlt, lambda = lambda))
      big_h_lambda <- t(apply(theta, 1, h_lambda, lambda = lambda, y = y))
      big_h_lambda_rep <- matrix(rep(big_h_lambda, length(lambda)),
                                 nrow = nrow(theta),
                                 ncol = length(lambda)
      )
      big_c <- matrix(rep(c, nrow(theta)),
                      ncol = length(c),
                      nrow = nrow(theta),
                      byrow = TRUE
      )
      return(colMeans(big_delta_lqlt * (big_h_lambda_rep - c)))
    }
    # nabla_LB(lambda, theta)
    
    # Now the control variate
    control_var <- function(lambda, theta) {
      # Get delta log q
      big_delta_lqlt <- t(apply(theta, 1, delta_lqlt, lambda = lambda))
      # Get h_lambda
      big_h_lambda <- t(apply(theta, 1, h_lambda, lambda = lambda, y = y))
      # Make it bigger
      big_h_lambda_rep <- matrix(rep(big_h_lambda, length(lambda)),
                                 nrow = nrow(theta),
                                 ncol = length(lambda)
      )
      # Now get delta log q times h
      big_nabla_log_q_h <- big_delta_lqlt * big_h_lambda_rep
      # Return the diagonals of the covariance and scale
      return(diag(cov(big_nabla_log_q_h, big_delta_lqlt)) / apply(big_delta_lqlt, 2, var))
    }
    # c <- control_var(lambda, theta)
    
    # LB estimate
    LB_lambda <- function(lambda, theta) {
      mean(apply(theta, 1, h_lambda, lambda = lambda))
    }
    # LB_lambda(lambda, theta)
    
    # Empirical version of derivative - comment this out if you want to create it
    # properly
    delta_lqlt <- function(lambda, theta, eps = 0.001) {
      k <- length(lambda)
      ans <- rep(NA, k)
      for (i in 1:k) {
        d <- rep(0, k)
        d[i] <- eps
        ans[i] <- (log_q(lambda + d, theta) - log_q(lambda - d, theta)) / (2 * max(d))
      }
      return(ans)
    }
    # K <- ncol(B)
    # lambda = c(rep(0, K), rep(1, K), 1, 1, 1, 1)
    # theta <- sim_theta(50, lambda)
    # delta_lqlt(lambda, theta[1,])
    
    # Natural
    
    # Run the VB function
    run_VB <- function(lambda, # Starting value of lambda
                       S = 100, # Number of samples to take
                       P = 10, # Maximum patience before you stop
                       beta_1 = 0.9, # Learning rates
                       beta_2 = 0.9, # Learning rates
                       tau = 1000, # Iteration at which learning rate starts to decrease
                       eps_0 = 0.1, # Raw learning rate multiplier
                       t_W = 50 # Time window for working out convergence
    ) {
      
      # Starting
      theta <- sim_theta(S, lambda)
      c <- control_var(lambda, theta)
      g_0 <- nabla_LB(lambda, theta)
      nu_0 <- g_0^2
      g_bar <- g_0
      nu_bar <- nu_0
      
      # Set up
      t <- 1
      patience <- 0
      stop <- FALSE
      LB <- rep(NA, t_W)
      max_LB_bar <- -Inf
      
      while (!stop) {
        if (t %% 10 == 0) print(t)
        
        # Generate new samples
        theta <- sim_theta(S, lambda)
        
        # Compute g_t
        g_t <- nabla_LB(lambda, theta, c)
        
        # Compute new control variate
        c <- control_var(lambda, theta)
        
        # Update the learning rates
        nu_t <- g_t^2
        g_bar <- beta_1 * g_bar + (1 - beta_1) * g_t
        nu_bar <- beta_2 * nu_bar + (1 - beta_2) * nu_t
        
        # Update the learning rate
        alpha_t <- min(eps_0, eps_0 * tau / t)
        
        # Update lambda
        lambda <- lambda + alpha_t * g_bar / sqrt(nu_bar)
        
        # Compute the moving average LB if out of warm-up
        if (t <= t_W) {
          # Compute a new lower bound estimate
          LB[t] <- LB_lambda(lambda, theta)
        } else {
          LB[1:(t_W - 1)] <- LB[2:t_W]
          LB[t_W] <- LB_lambda(lambda, theta)
          LB_bar <- mean(LB)
          max_LB_bar <- max(max_LB_bar, LB_bar)
          if (LB_bar >= max_LB_bar) {
            patience <- 0
          } else {
            patience <- patience + 1
          }
        }
        
        if (patience > P) {
          print("Completed!")
          stop <- TRUE
        }
        t <- t + 1
      }
      return(lambda)
    }
    
    rMVNormC <- function(n, mu, U){
      p <- length(mu)
      Z <- matrix(rnorm(p*n), p, n)
      # U <- chol(Omega) # By default R's chol fxn returns upper cholesky factor
      X <- backsolve(U, Z) # more efficient and stable than actually inverting
      X <- sweep(X, 1, mu, FUN=`+`)
      return(X)
    }
    
    
    # Now run it!
    
    lambdares[,i] <- run_VB(lambda = c(rep(0, K), rep(1, (((K * (K + 1)) / 2) + n_tracers * 2))))
    
  }
  
  sim_theta = function(S, lambda) {
    
    # For K parameters you will have
    # lambda is of length K+K*(K+1)/2 +n_tracers*2
    # mean <- lambda[1:K]
    # chol_prec is made up of lambda[(K + 1):(K+(K*(K+1))/2)]
    # Tau is made up of lambda[((K+(K*(K+1))/2)+1):((K+(K*(K+1))/2)+n_tracers*2)]
    # (f) ~ MVN(lambda[1:K], solve(crossprod(chol_prec)))
    
    
    mean <- lambda[1:K]
    # K*(K-1) precision terms
    chol_prec <- matrix(0, nrow = K, ncol = K)
    chol_prec[upper.tri(chol_prec, diag = TRUE)] <- lambda[(K + 1):(K + (K * (K + 1)) / 2)]
    
    # This is a placeholder for more advanced code using chol_prec directly
    theta <- cbind(
      t(rMVNormC(S, mu = mean, U = chol_prec)),
      matrix(rgamma(S * n_tracers,
                    shape = lambda[((K + (K * (K + 1)) / 2) + 1):(((K + (K * (K + 1)) / 2)) + n_tracers)],
                    rate = lambda[(((K + (K * (K + 1)) / 2)) + n_tracers + 1):(((K + (K * (K + 1)) / 2)) + n_tracers * 2)]
      ),
      nrow = S,
      ncol = n_tracers,
      byrow = TRUE
      )
    )
    
    return(theta)
  }
  
  rMVNormC <- function(n, mu, U){
    p <- length(mu)
    Z <- matrix(rnorm(p*n), p, n)
    # U <- chol(Omega) # By default R's chol fxn returns upper cholesky factor
    X <- backsolve(U, Z) # more efficient and stable than actually inverting
    X <- sweep(X, 1, mu, FUN=`+`)
    return(X)
  }
  
  mylist<-list(lambda = lambdares,
               n_sources = simmr_in$n_sources, 
               n_tracers = simmr_in$n_tracers,
               group = simmr_in$n_groups, 
               source_names = simmr_in$source_names,
               theta = sim_theta(3600, lambdares)
               )
  
  class(mylist) <- "simmr_ffvb_output"
  return(mylist)
}

#Check
#simmr_out = simmr_ffvb(simmr_in)





#I think we also need a different summary function?
summary_ffvb<-function(simmr_out, type = c("quantiles", "statistics", "correlations"), group = 1, ...){
  
  # Get the specified type
  type <- match.arg(type, several.ok = TRUE)
  K = simmr_out$n_sources
  n_tracers = simmr_out$n_tracers
  lambda = simmr_out$lambda
  n_groups = simmr_out$group
  names = simmr_out$source_names
  theta = simmr_out$theta
  

  
  
  for (i in 1:n_groups) {
    cat(paste("\nSummary for group", i))
    
    all_VB <- theta# Starting value of lambda
    
    p_fun <- function(x) exp(x)/sum(exp(x))
    p_VB <- t(apply(all_VB[,1:K], 1, p_fun))
    out_all <- matrix(cbind(p_VB, 1/sqrt(all_VB[,(K+1):(K+n_tracers)])), ncol = (K+n_tracers))
    
    out_quantiles <- array(rep(NA, 5*ncol(all_VB)*n_groups), dim = c(ncol(all_VB), 5, n_groups))
    #5 because 5 different quantiles taken
    
    colnames(out_quantiles) = c("0.025", "0.25", "0.5", "0.75", "0.975")
    rownames(out_quantiles)=c(simmr_out$source_names, rep("sd", n_tracers))

    
    out_statistics <- array(rep(NA,n_groups*2*ncol(all_VB)), dim = c(ncol(all_VB), 2, n_groups))
    colnames(out_statistics) = c("mean", "sd")
    rownames(out_statistics)=c(simmr_out$source_names, rep("sd", n_tracers))
 
    
    out_cor <- matrix(rep(NA,n_groups^2), ncol = n_groups, nrow = n_groups)
    
    out_quantiles[,,i] <- t(apply(out_all, 2, "quantile", probs = c(0.025, 0.25, 0.5, 0.75, 0.975)))
    #  coda:::summary.mcmc.list(object$output)$quantiles
    out_statistics[,,i] <- t(apply(out_all, 2, function(x) {
      return(c(mean = mean(x), sd = stats::sd(x)))
    }))
    # coda:::summary.mcmc.list(object$output)$statistics[,1:2]
    out_cor <- stats::cor(out_all)
    rownames(out_cor)=c(simmr_out$source_names, rep("sd", n_tracers))
    colnames(out_cor)=c(simmr_out$source_names, rep("sd", n_tracers))
    
    
    if ("quantiles" %in% type) {
      # Print out quantiles argument
      cat(paste0("\nquantiles\n"))
      print(round(out_quantiles[,,i], 3))
    }
    
    if ("statistics" %in% type) {
      cat(paste0("\nstatistics\n"))
      
      # Print out quantiles argument
      print(round(out_statistics[,,i], 3))
    }
    
    
  }
  
  if ("correlations" %in% type) {
    cat(paste0("\ncorrelations\n"))
    # Print out quantiles argument
    print(round(out_cor, 3))
  }
}

#Check
#summary_ffvb(simmr_out, type = c("quantiles", "statistics", "correlations"))



```

mix 1
```{r}
mix = matrix(c(-10.13, -10.72, -11.39, -11.18, -10.81, -10.7, -10.54, 
               -10.48, -9.93, -9.37, 11.59, 11.01, 10.59, 10.97, 11.52, 11.89, 
               11.73, 10.89, 11.05, 12.3), ncol=2, nrow=10)
colnames(mix) = c('d13C','d15N')
s_names = c("Zostera", "Grass", "U.lactuca", "Enteromorpha")
s_means = matrix(c(-14, -15.1, -11.03, -14.44, 3.06, 7.05, 13.72, 5.96), ncol=2, nrow=4)
s_sds = matrix(c(0.48, 0.38, 0.48, 0.43, 0.46, 0.39, 0.42, 0.48), ncol=2, nrow=4)
c_means = matrix(c(2.63, 1.59, 3.41, 3.04, 3.28, 2.34, 2.14, 2.36), ncol=2, nrow=4)
c_sds = matrix(c(0.41, 0.44, 0.34, 0.46, 0.46, 0.48, 0.46, 0.66), ncol=2, nrow=4)
conc = matrix(c(0.02, 0.1, 0.12, 0.04, 0.02, 0.1, 0.09, 0.05), ncol=2, nrow=4)
```

```{r}
simmr_in = simmr::simmr_load(mixtures=mix,
                     source_names=s_names,
                     source_means=s_means,
                     source_sds=s_sds,
                     correction_means=c_means,
                     correction_sds=c_sds,
                     concentration_means = conc)
```


```{r}
simmr_out = simmr_ffvb(simmr_in)
summary_ffvb(simmr_out)

n<-3600
all_vb <- simmr_out$theta
K = length(s_names)
f_VB <- all_vb[,1:K]


# Try the p 
#this gives what we want on the diag of the matrix plot
#Also the density plot
library(tidyverse)
p_fun <- function(x) exp(x)/sum(exp(x))
p_VB <- t(apply(all_vb[,1:K], 1, p_fun))
colnames(p_VB) <-simmr_out$source_names
for (i in 1:K) {
  print(data.frame(
    p = c( p_VB[,i]),
    Fit = c(c(rep("VB", n)))
  ) %>% ggplot(aes(x = p, fill = Fit)) + geom_density(alpha = 0.5) + ylim(0, 10)+
    ggtitle(paste("p: source",simmr_out$source_names[i]))) 
}

```

##Multiple groups

```{r}
mix = matrix(c(-11.36, -11.88, -10.6, -11.25, -11.66, -10.41, 
               -10.88, -14.73, -11.52, -15.89, -14.79, -17.64, -16.97, -17.25, 
               -14.77, -15.67, -15.34, -15.53, -17.27, -15.63, -15.94, -14.88, 
               -15.9, -17.11, -14.93, -16.26, -17.5, -16.37, -15.21, -15.43, 
               -16.54, -15, -16.41, -15.09, -18.06, -16.27, -15.08, -14.39, 
               -21.45, -22.52, -21.25, -21.84, -22.51, -21.97, -20.23, -21.64, 
               -22.49, -21.91, -21.65, -21.37, -22.9, -21.13, -19.33, -20.29, 
               -20.56, -20.87, -21.07, -21.69, -21.17, -21.74, -22.69, -21.06, 
               -20.42, -21.5, -20.15, -21.99, -22.3, -21.71, -22.48, -21.86, 
               -21.68, -20.97, -21.91, -19.05, -22.78, -22.36, -22.46, -21.52, 
               -21.84, -21.3, -21.39, -22.1, -21.59, -20.14, -20.67, -20.31, 
               -20.07, -21.2, -20.44, -22.06, -22.05, -21.44, -21.93, -22.47, 
               -22.27, -22.19, -22.81, -20.48, -22.47, -18.06, -20.72, -20.97, 
               -19.11, -18.4, -20.45, -21.2, -19.74, -20.48, -21.48, -17.81, 
               -19.77, -22.56, -14.72, -12.21, -12.35, -13.88, -14.43, -14.65, 
               -13.9, -14.12, -10.88, -10.44, -15.33, -13.78, -13.98, -15.22, 
               -15.25, -15.76, -15.78, -15.49, -13.02, -15.3, -15.55, -14.35, 
               -14.99, -14.83, -16.18, -15.01, -12.87, -14.67, -13.84, -14.89, 
               -13.33, -15.04, -14.29, -15.62, -13.99, -15.06, -15.06, -15, 
               -14.55, -13.32, -14.34, -14.47, -14.31, -14.18, -16.18, -16.25, 
               -15.92, -15.35, -14.29, -15.92, -15.35, -20.22, -21.4, -19.97, 
               -20.78, -20.61, -20.58, -20.19, -20.71, -20.59, -20.09, -19.37, 
               -20.41, -20.84, -20.75, -20.29, -20.89, -19.69, -20.41, -21.24, 
               -19.33, -25.87, -25.4, -27.23, -27.52, -24.55, -17.36, -24.7, 
               -27.76, -28.92, -25.98, -26.77, -28.76, -27.7, -24.75, -25.47, 
               -26.58, -28.94, -29.13, -26.65, -28.04, -27.5, -29.28, -27.85, 
               -27.41, -27.57, -29.06, -25.98, -28.21, -25.27, -14.43, -27.4, 
               -27.76, -28.45, -27.35, -28.83, -29.39, -28.86, -28.61, -29.27, 
               -20.32, -28.21, -26.3, -28.27, -27.75, -28.55, -27.38, -29.13, 
               -28.66, -29.02, -26.04, -26.06, -28.52, -28.51, -27.93, -29.07, 
               -28.41, -26.42, -27.71, -27.75, -24.28, -28.43, -25.94, -28, 
               -28.59, -22.61, -27.34, -27.35, -29.14, 10.22, 10.37, 10.44, 10.52, 10.19, 10.45, 9.91, 11.27, 
               9.34, 11.68, 12.29, 11.04, 11.46, 11.73, 12.29, 11.79, 11.49, 
               11.73, 11.1, 11.36, 12.19, 11.03, 11.21, 10.58, 11.61, 12.16, 
               10.7, 11.47, 12.07, 11.75, 11.86, 12.33, 12.36, 11.13, 10.92, 
               12.42, 10.95, 12.28, 11.04, 10.76, 10.99, 10.78, 11.07, 10.2, 
               11.67, 7.53, 10.65, 10.58, 11.13, 7.73, 10.79, 10.47, 10.82, 
               10.41, 11.1, 10.95, 10.76, 10.83, 10.25, 10.52, 9.94, 9.94, 11.61, 
               10.65, 10.76, 11.11, 10.2, 11.27, 10.21, 10.88, 11.21, 11.36, 
               10.75, 12.38, 11.16, 11.57, 10.79, 11.13, 10.72, 10.99, 10.38, 
               10.95, 10.75, 10.75, 11.05, 10.66, 10.61, 10.9, 11.14, 10.33, 
               10.83, 10.75, 9.18, 9.03, 9.05, 8.6, 8.29, 10.32, 10.28, 6.47, 
               11.36, 10.75, 11.13, 11.37, 10.86, 10.54, 10.39, 10.66, 9.99, 
               11.65, 11.02, 10.67, 8.15, 11.12, 10.95, 11.2, 10.76, 11.32, 
               10.85, 11.74, 10.46, 10.93, 12.3, 10.67, 11.51, 10.56, 12.51, 
               13.51, 11.98, 12.2, 10.48, 12.4, 13, 11.36, 12.08, 12.39, 12.28, 
               12.6, 11.3, 11.1, 11.42, 11.49, 12, 13.35, 11.97, 13.35, 12.75, 
               12.55, 12.3, 12.51, 12.61, 10.98, 11.82, 12.27, 12.11, 12.11, 
               12.89, 12.99, 12.29, 11.89, 12.74, 12.29, 11.89, 10.56, 9.27, 
               10.54, 10.97, 10.46, 10.56, 10.86, 10.9, 11.06, 10.76, 10.64, 
               10.94, 10.85, 10.45, 11.15, 11.23, 11.16, 10.94, 11.2, 10.71, 
               9.55, 8.6, 9.67, 8.17, 9.81, 10.94, 9.49, 9.46, 7.94, 9.77, 8.07, 
               8.39, 8.95, 9.83, 8.51, 8.86, 7.93, 8, 8.33, 8, 9.39, 8.01, 7.59, 
               8.26, 9.49, 8.23, 9.1, 8.21, 9.59, 9.37, 9.47, 8.6, 8.23, 8.39, 
               8.24, 8.34, 8.36, 7.22, 7.13, 10.64, 8.06, 8.22, 8.92, 9.35, 
               7.32, 7.66, 8.09, 7.3, 7.33, 7.33, 7.36, 7.49, 8.07, 8.84, 7.93, 
               7.94, 8.74, 8.26, 9.63, 8.85, 7.55, 10.05, 8.23, 7.74, 9.12, 
               7.33, 7.54, 8.8), ncol=2, nrow=251)
colnames(mix) = c('d13C','d15N')
s_names = c("Zostera", "Grass", "U.lactuca", "Enteromorpha")
s_means = matrix(c(-11.17, -30.88, -11.17, 
                   -14.06, 6.49, 4.43, 11.19, 9.82), ncol=2, nrow=4)
s_sds = matrix(c(1.21, 0.64, 1.96, 1.17, 1.46, 2.27, 1.11, 0.83), ncol=2, nrow=4)
c_means = matrix(c(1.63, 1.63, 1.63, 1.63, 3.54, 3.54, 3.54, 3.54), ncol=2, nrow=4)
c_sds = matrix(c(0.63, 0.63, 0.63, 0.63, 0.74, 0.74, 0.74, 0.74), ncol=2, nrow=4)
conc = matrix(c(0.36, 0.4, 0.21, 0.18, 0.03, 0.04, 0.02, 0.01), ncol=2, nrow=4)
grp = as.integer(c(1, 1, 1, 1, 1, 1, 1, 1, 1, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 
        2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 3, 3, 3, 
        3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 
        3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 
        3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 
        3, 3, 3, 3, 3, 3, 3, 3, 4, 4, 4, 4, 4, 4, 4, 4, 4, 4, 5, 5, 5, 
        5, 5, 5, 5, 5, 5, 5, 5, 5, 5, 5, 5, 5, 5, 5, 5, 5, 5, 5, 5, 5, 
        5, 5, 5, 5, 5, 5, 5, 5, 5, 5, 5, 5, 5, 5, 5, 5, 5, 6, 6, 6, 6, 
        6, 6, 6, 6, 6, 6, 6, 6, 6, 6, 6, 6, 6, 6, 6, 6, 7, 7, 7, 7, 7, 
        7, 7, 7, 7, 7, 7, 7, 7, 7, 7, 7, 7, 7, 7, 7, 7, 7, 7, 7, 7, 7, 
        7, 7, 7, 7, 7, 7, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 
        8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8))
```

Note that the `grp` object above is specified to be an integer but can also be a factor, the levels of which will appear in plots. However, when specifying groups in later commands you should use the integer values to reference which groups to plot

Next it is a matter of following the simmr commands as before to load in, with an extra argument specifying the groups:

```{r}
simmr_groups = simmr::simmr_load(mixtures=mix,
                     source_names=s_names,
                     source_means=s_means,
                     source_sds=s_sds,
                     correction_means=c_means,
                     correction_sds=c_sds,
                     concentration_means = conc,
                     group=grp)
```

```{r}
simmr_groups_out = simmr_ffvb(simmr_groups)
summary_ffvb(simmr_groups_out)
```


##One isotope

```{r}
mix = matrix(c(-10.13, -10.72, -11.39, -11.18, -10.81, -10.7, -10.54, 
               -10.48, -9.93, -9.37), ncol=1, nrow=10)
colnames(mix) = c('d13C')
s_names = c("Zostera", "Grass", "U.lactuca", "Enteromorpha")
s_means = matrix(c(-14, -15.1, -11.03, -14.44), ncol=1, nrow=4)
s_sds = matrix(c(0.48, 0.38, 0.48, 0.43), ncol=1, nrow=4)
c_means = matrix(c(2.63, 1.59, 3.41, 3.04), ncol=1, nrow=4)
c_sds = matrix(c(0.41, 0.44, 0.34, 0.46), ncol=1, nrow=4)
conc = matrix(c(0.02, 0.1, 0.12, 0.04), ncol=1, nrow=4)
```

Now load in with `simmr_load`:
```{r}
simmr_in_1D = simmr::simmr_load(mixtures=mix,
                     source_names=s_names,
                     source_means=s_means,
                     source_sds=s_sds,
                     correction_means=c_means,
                     correction_sds=c_sds,
                     concentration_means = conc)
```

Create a plot. `plot.simmr_input` automatically creates a 1D version of these plots:
```{r,results='hide'}
plot(simmr_in_1D)
```

Now run simmr:

```{r}
simmr_run_1D = simmr_ffvb(simmr_in_1D)
summary_ffvb(simmr_run_1D)
```

